\Large{Adversarial Variational Optimization of Non-Differentiable Simulators}

### 2017-06-22 paper.tex
save

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log q_\psi(\mathbf{c})$ of the proposal
is known and that one can evaluate $f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then
one can construct empirical estimates of Eqn.~\ref{eqn:approx-grad}, which can
in turn be used to minimize $U(\psi)$ with stochastic gradient descent (or a variant thereof, like Adam~\cite{2014arXiv1412.6980K} to account for scaling invariance and robustness to noisy gradients).

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ and $\tilde{\nabla}_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
% That is,
% \begin{equation}
%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
% \end{equation} which we can approximate with mini-batches of
% generated data
% \begin{equation}
%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
% \end{equation}
% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP.
Let us note however that the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.

>>>>

Under mild restrictions outlined in  \citep{2012arXiv1212.4507S}, the bound $U(\psi)$ is differentiable, and using the log-likelihood trick it comes:
\begin{align}\label{eqn:approx-grad}
    \nabla_\psi U(\psi) &= \nabla_\psi \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c})] \nonumber \\
    &= \nabla_\psi \int f(\mathbf{c})  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int f(\mathbf{c}) \nabla_\psi q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \int \left[ f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c}) \right]  q_\psi(\mathbf{c})  d\mathbf{c} \nonumber \\
    &= \mathbb{E}_{\mathbf{c} \sim q_\psi(\mathbf{c})} [f(\mathbf{c}) \nabla_\psi \log q_\psi(\mathbf{c})]
\end{align}
Effectively, this means that provided that the score function $\nabla_\psi \log
q_\psi(\mathbf{c})$ of the proposal is known and that one can evaluate
$f(\mathbf{\mathbf{c}})$ for any $\mathbf{c}$, then one can construct empirical
estimates of Eqn.~\ref{eqn:approx-grad}, which can in turn be used to minimize
$U(\psi)$ with stochastic gradient descent (or a variant thereof, like
Adam~\cite{2014arXiv1412.6980K} or the Natural Evolution Strategy
algorithm~\citep{2011arXiv1106.4487W}, to account for scaling invariance and
robustness to noisy gradients).

@@

In this work, we propose to rely on variational optimization to minimize ${\cal
L}_d$ and ${\cal L}_g$, thereby bypassing the non-differentiability of $g$. More
specifically, we consider a proposal distribution $q_\psi(\theta)$ over the
parameters of $g$ and minimize in alternance the variational upper bounds
\begin{align}
    U_d &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_d ] \label{eqn:vo-ud} \\
    U_g &= \mathbb{E}_{\theta \sim q_\psi} [ {\cal L}_g ] \label{eqn:vo-ug}
\end{align} respectively over $\phi$ and $\psi$.
When updating
$d$, unbiased gradient estimates of $\nabla_\phi U_d$ can be obtained by
evaluating the exact and known gradient of $U_d$ over mini-batches of true and
generated data, as ordinarily done in stochastic gradient descent. When updating
$g$, estimates of $\nabla_\psi U_g$ can be derived with forward
simulations, as described in the previous section.
% That is,
% \begin{equation}
%     \nabla_\psi U_g = \mathbb{E}_{\theta \sim q_\psi(\theta), \mathbf{z} \sim p_z}  [-d(g(\mathbf{z};\theta);\phi) \nabla_\psi \log q_\psi(\theta)],
% \end{equation} which we can approximate with mini-batches of
% generated data
% \begin{equation}
%     \nabla_\psi U_g \approx \frac{1}{M} \sum_{m=1}^M -d(g(\mathbf{z}_m; \theta_m); \phi) \nabla_\psi \log q_\psi(\theta_m)
% \end{equation}
% for $\mathbf{z}_m \sim p_z$ and $\theta_m \sim q_\psi$.
\glnote{Can we exploit the fact that $\nabla_\mathbf{x} d(\mathbf{x})$
is known exactly for building better estimates $\tilde{\nabla}_\psi U_g$?
% In practice, using
% \begin{equation}
%     \tilde{\nabla}_\psi U_g = \frac{1}{M} \sum_{m=1}^M g(\mathbf{z}_m; \theta_m) \nabla_\psi \log q_\psi(\theta_m) \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}_m; \phi))
% \end{equation}
% works as good it seems, but I am not sure to precisely understand why, nor whether this is correct...
% It feels like this is approximating the chain rule $\nabla_\theta g(\theta)  \nabla_{\tilde{\mathbf{x}}} (-d(\tilde{\mathbf{x}}))$.
}

@@

For completeness, Algorithm~\ref{alg:avo} finally outlines the full adversarial variational
optimization procedure, as built on top of WGAN-GP.
Let us note that the variational relaxation could similarly be coupled with
other variants of GANs and/or of evolution strategies.
